---
title: "viz_and_eda_day1"
output: html_document
date: "2025-09-25"
---

```{r}
library(tidyverse)
library(ggridges)
library(p8105.datasets)
```

Import the weather data

```{r}
data("weather_df")
```

Making our first plot :-)

Use the `+` operator to link together your code in ggplot (kind of like the pipe operator but for ggplot)

```{r}
ggplot(data = weather_df, mapping = aes(x = tmin, y = tmax)) + #telling R which data, and which variables
  geom_point()

```

Another way to do the same thing:

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()
```

This is a way to save the plot to your environment (assign it to an object):

```{r}
ggp_weather_scatterplot =
  weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()

ggp_weather_scatterplot #this is how I can see what I just saved
```

Check that some rows are missing.

```{r}

weather_df |> 
  filter(is.na(tmax)) #this will list out in the console all observations with missing tmax

```

## Fancier scatterplots!

alpha ranges from 0-1 (1=totally opaque, 0=totally see-through)
```{r}

weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) + #different color for each name
  geom_point(alpha = 0.3) +#changes opacity of points so you can see where the highest density of data lies
  geom_smooth(se = FALSE) #puts a smooth line of best fit (for each color); se = FALSE removes SE bar

```


Where you define aesthetics can matter

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = 0.3, size = 0.6) + 
  #makes my points a different size (<1 = smaller, >1 = bigger)
  geom_smooth(se = FALSE) 
#difference here is I get a best fit line for the whole plot, not for each name, because the color statement was not at the top, therefore it was not inherited by every layer below
```

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name, size = prcp), alpha = 0.3) + #bigger points w/ more rain, smaller w/ less rain
  geom_smooth(se = FALSE)
```

Use faceting real quick

```{r}

weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = 0.3, size = 0.8) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name) #before squiggle `~` = row, after ~ = column --> separates into a grid of multiple plots

weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = 0.3, size = 0.8) +
  geom_smooth(se = FALSE) + 
  facet_grid(name ~ .) #3 rows of plots instead of 3 columns of plots

```






